Visual Diagnostics of a Model Explainer -- Tools for the Assessment of LIME Explanations

### 2020-06-04 paper.Rnw
edits through 3.0

The concept of LIME is relatively simple: use an interpretable model to approximate a complex model in a local region. However, a practical implementation of LIME is not straightforward, and there is research is being done to improve the procedure \citep{laugel:2018}. The current implementations of LIME\footnote{https://github.com/thomasp85/lime} \footnote{https://github.com/goodekat/lime} offer various tuning parameters the user specifies when applying LIME (see Section \ref{background}) that affect the explainer model and ultimately, the explanation. Since the explainer model is an approximation of the complex model and not a direct interpretation, the explanations produced by an explainer model are subject to the quality of the approximation. Thus, in order to achieve accurate explanations, the tuning parameters selected need to be assessed.

@@

Several sources in the literature discuss the performance of LIME. One of the biggest difficulties with LIME is determining how to specify a local region \citep{molnar:2019} \citet{laugel:2018}. This is due to an unclear definition of what a local region means and how to apply LIME to achieve an appropriate local region as demonstrated by \autoref{fig:concept-plot-bad}. \citet{alvarezmelis:2018} raise a concern pertaining to the robustness of explanations from LIME and other model explainers: they find that even small changes in predictor variables can lead to very different LIME explanations. Additionally, \citet{ribeiro:2016} acknowledge that if a linear models is used as the explainer, LIME relies on a linear approximation of the explainer model to the complex model and state ``if the underlying model is highly non-linear even in the locality of the prediction, there may not be a faithful explanation".

@@

LIME was introduced in 2016 by \citet{ribeiro:2016}. The original authors provide an implementation of LIME in a Python package\footnote{https://github.com/marcotcr/lime}. An adaption of the Python package in R is by Thomas Lin-Pedersen\footnote{https://github.com/thomasp85/lime}. In this paper, any details provided about the implementation of LIME are based on the R package.

@@

@
\caption{Plots of $x_2$ versus $x_1$ from the training (left) and testing (right) set of the \data \ introduced in Section \ref{procedure}. The true classification boundary is shown as the solid black line in both plots. The color of the training data points represents the value of the observed response variable ($y$). The color of the testing data points represents the random forest probability that an observation belongs to the category of blue ($\hat{y}$). The \Sexpr{dim(sine_data_test %>% filter(y != rfpred))[1]} cases that are misclassified by the random forest are identified by black circles. The prediction of interest to explain is indicated by a diamond.}
\label{fig:sine-plot-data}
\end{figure*}

@@

\kgc{This is where I left off with the editing:} We apply LIME to the \data \ using \emph{lime} with the default settings. The user is asked to specify the number of features shown in the explanations. Here, we set this value to two, because we expect that $x_1$ and $x_2$ will be the main drivers in the prediction. Note that the complexity of an explanation goes hand in hand with the number of features selected to be included in the explanation. Setting the number of features to a small number will make the explanation simpler but not necessarily more correct.

To demonstrate the proposed diagnostic plots, we applied six different implementations of LIME to the sine data to select the top two features associated with each of the random forest predictions in the testing dataset. This was done using six different sets of tuning parameters via \sout{our} R package \emph{limeaid} \footnote{Available on GitHub at https://github.com/goodekat/limeaid}. \kge{(In this paper, we use a forked version\footnote{https://github.com/goodekat/lime} of the development version of the R package \emph{lime} (\Sexpr{packageVersion("lime")}) \citep{pedersen:2020}. The forked version of \emph{lime} allows us to export internal values of the procedure not already exported by \emph{lime}.)} Quantile bins were used for five of the implementations, and the number of bins for these implementations was varied from 2 to 6. The sixth implementation of LIME used the kernel density estimates. The feature selection method and the computation of the weights were set to the default options in \emph{lime} of forward selection and Gower distance raised to an exponent of 1.

@@

\begin{figure*}[!htp]
<<sine-plot1, out.width = '\\textwidth', fig.width = 10, fig.height = 4, warning = FALSE>>=

@@

@@

\begin{figure}[!t]
<<sine-feature-heatmap, out.width = '0.5\\textwidth', fig.width = 5, fig.height = 5.5>>=
feature_heatmap(sine_lime_explain$explain,
                order_method = "PCA") +
  theme_bw(base_family = "Times", base_size = 10) +
  theme(legend.position = "bottom",
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank()) +
  labs(y = "Case")
@
\caption{Feature heatmap of the various LIME implementations applied to the \data. The cases from the test set are plotted on the y-axis, and the implementation methods are included on the x-axis. The colors of the tiles indicate the feature selected by LIME for the corresponding case and implementation method. The top faceted row includes the first features selected by LIME, and the bottom faceted row includes the second features selected by LIME. The faceted columns separate the kernel density implementation method from the quantile bin implementations. The vertical striping seen in the plot indicates that LIME is not consistent in selecting features across implementation methods.}
\label{fig:sine-feature-heatmap}
\end{figure}

>>>>

The concept of LIME is relatively simple: use an interpretable model to approximate a complex model in a local region. However, a practical implementation of LIME is not straightforward, and there is research is being done to improve the procedure \citep{laugel:2018}. The current implementations of LIME\footnote{https://github.com/marcotcr/lime} \citep{pedersen:2020} offer various tuning parameters the user specifies when applying LIME (see Section \ref{background}) that affect the explainer model and ultimately, the explanation. Since the explainer model is an approximation of the complex model and not a direct interpretation, the explanations produced by an explainer model are subject to the quality of the approximation. Thus, in order to achieve accurate explanations, the tuning parameters selected need to be assessed.

@@

Several sources in the literature discuss the performance of LIME. One of the biggest difficulties with LIME is determining how to specify a local region \citep{laugel:2018} \citep{molnar:2019}. This is due to an unclear definition of a "local region" and how to apply LIME to achieve an appropriate local region as demonstrated by \autoref{fig:concept-plot-bad}. \citet{alvarezmelis:2018} raise a concern pertaining to the robustness of explanations from LIME and other model explainers: they find that even small changes in predictor variables can lead to very different LIME explanations. Additionally, \citet{ribeiro:2016} acknowledge that if a linear models is used as the explainer, LIME relies on a linear approximation of the explainer model to the complex model and state ``if the underlying model is highly non-linear even in the locality of the prediction, there may not be a faithful explanation".

@@

LIME was introduced in 2016 by \citet{ribeiro:2016}. The original authors provide an implementation of LIME in a Python package\footnote{https://github.com/marcotcr/lime}. An adaption of the Python package in R is by Thomas Lin-Pedersen \citep{pedersen:2020}. In this paper, any details provided about the implementation of LIME are based on the R package.

@@

@
\caption{Plots of $x_2$ versus $x_1$ from the training (left) and testing (right) sets of the \data \ introduced in Section \ref{diagnostics}. The true classification boundary is shown as the solid black line in both plots. The color of the training data points represents the value of the observed response variable ($y$). The color of the testing data points represents the random forest probability that an observation belongs to the category of blue ($\hat{y}$). The \Sexpr{dim(sine_data_test %>% filter(y != rfpred))[1]} cases that are misclassified by the random forest are identified by black circles. The prediction of interest to explain is indicated by a diamond.}
\label{fig:sine-plot-data}
\end{figure*}

@@

@@

\begin{figure*}[!tp]
<<sine-plot1, out.width = '\\textwidth', fig.width = 10, fig.height = 4, warning = FALSE>>=

@@

We perform six applications of LIME with different tuning parameters to the random forest predictions from all observations in the sine data test set. To do this, we use the R package \emph{limeaid} \footnote{Available on GitHub at https://github.com/goodekat/limeaid} (\Sexpr{packageVersion("limeaid")}), which applies a forked version\footnote{https://github.com/goodekat/lime} of the development version of the R package \emph{lime} (\Sexpr{packageVersion("lime")}) \citep{pedersen:2020}. The forked version allows us to export internal values from \emph{lime}. The tuning parameters are as follows. For each application, we specify that LIME returns the top two features in the explanation, because we expect $x_1$ and $x_2$ to be the important features. Note that the complexity of an explanation goes hand in hand with the number of features selected to be included in the explanation. Setting the number of features to a small number will make the explanation simpler but not necessarily more correct. A quantile bin based simulation method is used for five of the implementations with the number of bins varying from 2 to 6 by application. The sixth implementation of LIME uses a kernel density simulation method. The default methods for feature selection (forward selection) and the computation of the weights (Gower distance raised to an exponent of 1) are used for all applications.

@@

\begin{figure}[!t]
<<sine-feature-heatmap, out.width = '0.5\\textwidth', fig.width = 4, fig.height = 4>>=
feature_heatmap(sine_lime_explain$explain,
                order_method = "PCA") +
  theme_bw(base_family = "Times", base_size = 10) +
  theme(legend.position = "bottom",
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank()) +
  labs(y = "Case")
@
\caption{Feature heatmap of the various LIME implementations applied to the \data. The cases from the test set are plotted on the y-axis, and the implementation methods are included on the x-axis. The colors of the tiles indicate the feature selected by LIME for the corresponding case and implementation method. The top faceted row includes the first features selected by LIME, and the bottom faceted row includes the second features selected by LIME. The faceted columns separate the kernel density implementation method from the quantile bin implementations. The vertical striping seen in the plot indicates that LIME is not consistent in selecting features across implementation methods.}
\label{fig:sine-feature-heatmap}
\end{figure}
